Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-05-28 Manuscript.Rmd
text edits

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.  That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag, [@Nijhawan1994]; or the flash grab effect, [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].

@@

The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory. Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.

The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus after having seen it (a retrace procedure). This will allow us to distinguish between a spatial limit to the illusion or a temporal limit.

@@

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, or the tip of the stylus.

@@

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ (or ~XX _dva_), using 1.5 phases of a sine wave in the full stimulus, so that each sine phase equals ~0.582 _cm_ or ~XX _dva_. By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ (~ XX _dva_) and was set so that stimulus would move from either extreme to the other (a motion we call a 'pass') in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ or roughly XX, XX or XX _dva/s_ in the middle of the screen. For simplicity, these will be reported as _seconds per pass_.
- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)

@@

[

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second. We then calculated the instantaneous heading (direction, disregarding velocity) of a smooth spline interpolated version of the tracking trajectories. This heading measure should depend on the illusion strength, and should change when a (spontaneous) reset is reported in the tracking data.

]

[

For comparison, we use the same heading measure in experiment 2, but looking at individual trajectories reveals more points that are potential spontaneous resets. Using a smooth spline interpolated version of the trajectory we then detect direction changes (beyond ~0.39 _cm_ from the starting point, as the initial movement is often jittery). These direction changes are taken as reset points. We use the 54 average coordinates of the reset points in all 9 participants and 6 stimulus conditions for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

]

@@

Two similar tasks were used in this experiment. In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 _cps_). Before stimulus onset

The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 _cm_ (~XXX _dva_) and 2 _s_ each. The initial external motion direction (forward or backward) was counterbalanced. When the external motion inverted at the far and close ends of the trajectory, the internal motion was also inverted. Each stimulus was repeated 8 times, with randomized position of the fixation point (left or right) of the stimulus and counterbalanced initial external motion direction (forward or backward). This task allowed testing how sensitive online tracking is to the illusion, and served to gauge what resets would look like in tracking data.

**DOUBLE PARAGRAPH; CHECK WHAT IS MISSING**

In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path (or 3 cycles).

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_=<.001). This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

In a second experiment, we used a retracing task where the participants drew what they saw of a single pass stimulus immediately after presentation. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. In this task, the double-drift gabor started at the near end of the display, and moved forward along the horizontal display away from the participant. The gabor could have one of two external speeds (corresponding to 3 or 4 _s_ presentation time), and one of three internal speeds (2, 3 or 4 _cps_). Apart from re-tracing the perceived paths, in half the blocks participants indicated the angle of the first part of the stimulus motion by orienting a line (75 _px_ = 3 _cm_) originating from the starting point of the gabor’s path. This task has been used before [XXXX]. In the meantime, Nakayama & Holcombe [-@Nakayama2019] have also used a re-tracing paradigm to study triggered resets.

@@

Participants alternated between blocks of two kinds of trials (in counterbalanced order). In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds. During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus. After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways. In one kind of block, the participant re-traced the perceived path of the gabor using the stylus. The drawn path was shown as a line on the screen. Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key. In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving. Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER. The arrow report task has been used before (ref?) and can be used as validation of this task. The re-tracing task has also been used before, but in order to test evoked resets [@Nakayama2019].

@@

[Actual results] For the majority of participants (X/9) we see that most trials show at least one reset (>90%?). In the remainder we still see >50% of the trials have a reset.

@@

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus continuously. While the direction of the tracking movements scaled with the internal motion of the stimulus (was sensitive to the strength of the illusion), there were no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion [just] as well, and in most trials there is a clear reset point. The positions of these reset points do no vary much with elapsed time, but are better explained as triggered when the distance between perceived and true position of the stimulus exceeds some threshold. This indicates that while the perceived location of a moving object can be heavily biased by its well-defined, internal motion, an uncertain (probably retinal?) position signal is still used to keep the perceived position of moving objects close to reality.

§3 (one computational model):

- Explain paper with model by Kwon, Tadin & Knill, and VSS poster with fat-tailed noise.

While this model produces reasonable resets, the properties of resets, such as their spatial or temporal distributions have not been characterized yet. An alternative explanation for resets could be that the true but imprecise position that is available for eye-movements can limit or reset the perceived position under some circumstances. In order to get at the properties of resets

If the double-drift illusion is taken as a vector average between the external motion and some proportion of the internal motion, the initial path of the stimulus can be modelled as a straight line from the origin at some angle. This straight line segment should intersect the position of the first reset. We will now see if resets occur at a given distance perpendicular to the path of the real stimulus, after a given amount of time, or perhaps a combination of the two.

This means that it seems that resets occur predominantly at some given discrepancy between the true and illusory position of the stimulus, with a small contribution of time. It may that after seeing the illusion for some time, a ‘source estimation’ mechanism can no longer account for both the perceived, illusory position and the imprecise retinal position, as coming from the same stimulus. Then the visual system changes the position of the stimulus to the position derived from retinotopic information (instead of opting for two separate sources, which is highly unlikely in this case), resulting in a “reset”.

@@

We find that online tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in single-pass retracing. The illusion appears to resets predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true and illusory position of the stimulus.

@@

The research was support by funding from NSERC Canada to PC and DYPH , as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).

>>>>

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet. While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets. That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to resets.

Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure). By investigating the time and location of resets in the recorded trajectories, we can distinguish between a spatial limit to the illusion or a temporal limit.

@@

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.

Participants were to be made For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assume the participant is close to the setup, with their eyes at the height of the monitor. Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are underestimating the size of stimuli.

@@

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ). The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions. The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_. By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In experiment 2, participants first watched a double-drift stimulus and then either reproduce the perceived path of the stimulus by retracing it on the tablet, or indicate the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below.

@@

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment). We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength. Moreoever, when spontaneous resets are either time-dependent or space-dependent, resets should occur at roughly the same points and in the same direction in a given condition for every participant. This means that spontaneous resets should result in systematic deviations of every participant's average heading over time.

For comparison, we use the same heading measure in experiment 2. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

@@

In the first task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

@@

In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occured. The reset points can then be used to determine if resets are time-limited or space-limited.

@@

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. Nakayama & Holcombe [-@Nakayama2019; -@Nakayama2019b] have also used a re-tracing paradigm, in their case to study triggered resets. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After, the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path. They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise. They could "key in" their response by pressing ENTER. In both tasks, there was no time-limit for the response, and careful responding was encouraged.

Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task.

@@

On average, the algorithm detects resets in 75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in one of the six conditions for any participant is 7 (out of 24), with the average at 17. Since participants experiencing a "hit-the-wall reset" could display a direction change in their paths only by chance, 50% of trials with resets seems reasonable. Neither percept is incorrect, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets.

@@

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online. While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.

There is a generative model, based on Bayesian cue integration, that can explain many phenomena with motion induced position shifts. This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatbile with a good portion of our data. The model posits there are three signals contributing to an object's perceived position: it's global motion, it's local motion (or texture motion) and it's position. Especially when the directly perceived position is a noisy or weak signal, the two motion signals contribute more, as in the double-drift or curveball illusion. In this stimulus, the model predicts that the sensory motion signals bias perceived position away from the true location, but at some point this is balanced by the weak, noisy sensory position signal that bias perceived position back towards the true location [@Kwon2015, Fig 3]. When this balance is reached the perceived path of the object would be parallel to but somewhat displaced from the true object path. This pattern of perceived paths is exhibited by some of our participants, and we have called these "hit-the-wall resets" here.

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or perhaps they do not employ any causal inference process.

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2019; @Nakayama2019b], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered. This should however be tested with eye-tracking data, so that this will remain a speculation here.

@@

We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in delayed, single-pass re-tracing. The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus.

@@

The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).
